---
title: "Advanced R C2"
author: "Phil"
output: html_document
---

##Vectors

- two flavours, atomic vectors and lists
- atomic vectors must be all the same type, lists can have elements of different types
- typeof() length and attributes()


### Atomic Vectors

```{r}
#can have integer, numeric, logical and
#character vectors (also complex and raw, but rare)

int_var <- c(1L, 6L, 10L)
typeof(int_var)
is.integer(int_var)
is.atomic(int_var)
is.list(int_var)
```

- note that is.numeric() tests for numberliness, and will give true for integer and double.

#### Coercion of atomic vectors

- if comine an atomic vector of different types, they will be coerced to the most flexible type

```{r}
str(c('a',1))

#useful that TRUE and FALSE get coerced to 0 and 1 - can use sum() or mean()
as.numeric(c(TRUE, FALSE, TRUE))
```

###Lists

- lists can have elements of any type, including lists. This makes them recursive vectors

```{r,}
#can use both list() or c()
x <- list(1,'a', TRUE, c(1,2,3))
y <- c(1, 'a', TRUE, c(1,2,3))

#c() will combine several lists at once. it will coerce vectors to list before combining
x <- list(list(1, 2), c(3, 4))
y <- c(list(1,2), c(3,4))
str(x)
str(y)
```
- coerce  to list with as.list(), check if a list with is.list()
- NA is a logical vector, as it will be coerced to any other type if put in a vector. c(FALSE, NA_character_) will coerce FALSE to a character


##Attributes

- all objects can have arbitraty additional attributes to store metadata
- can be accessed individuallu with attr() or all at once as a list with attributes()

```{r,}
x <- 1:10
attr(x, 'my_attribute') <- 'this is a vector'
attr(x, 'my_attribute')
attr(x, 'my_attribute1') <- 'my name is phil'

str(attributes(x))
```
- structure() is not the same as str(). structure() returns the given object with further attributes set

```{r}
structure(1:10, my_attribute='This is a vector')
```

- most attributes are lost when modifying vectors
- the only attributres not lost are names, dimension and class. these have specific accessor functions names(), class() and dim(). DO NOT attr(x, 'names'), for example.

###Names

- can name a vector:

```{r}
x <- c(a=1, b=2, c=3)
x <- 1:3 ; names(x) <- c('a', 'b', 'c')
x <- setNames(1:3, c('a', 'b', 'c'))
```
- however names are not essential: if all missing names() returns null, if some missing names() returns emty string for those elements

###Factors

- remember as(numeric(as.character(blah))) to convert from factor to numeric

```{r}
gender <- factor(c('m', 'm', 'm'), levels = c('m', 'f'))
table(gender)
```
