---
title: "C6"
output: html_document
---

## Function Components

* all functions have three parts: body() - the code inside the function
* the formals() - the list of agruments which controls how you call the function
* the envirmonment() - the 'map' of the functions variables


```{r}
f <- function(x) x^2
f
formals(f)
body(f)
environment(f)
```

* R functions can posess any number of additional attributes

* Primitave functions call C code directly, so formals(), body() and environment are NULL

```{r}
sum
formals(sum)
body(sum)
environment(sum)
```

## Lexical scoping

* lexical scoping looks up symbol values based upn how how functions were nested when they were created, not when they were called.

### Name masking

* most basic principle of lexical scoping

```{r}
f <- function(){
  x <- 1
  y <- 2
  c(x, y)
}

f()

rm(f)

# If name not defined in function, R will look one level up
x <- 1
f <- function(){
  y <- 2
  c(x, y)
}

f()
rm(f)

# same rules apply to functions inside functions

x <- 1
f <- function(){
  y <- 2
  i <- function(){
    z <- 3
    c(x, y, z)
  }
  i()
}

f()
rm(f)
```

### functions vs variables

* The principle applies regardless of type of associated value - finding functions works the same way as finding variables

```{r}
l <- function(x) x + 1
m <- function() {
  l <- function(x) x * 2
  l(10)
}
m()
```

* do not give variables and functions the same name... quirks will happen!

### a fresh start

```{r}
j <- function() {
  if (!exists("a")){
    a <- 1
  } else {
    a <- a + 1
  }
  print(a)
}
j()
```

* every time the function is called, it returns 1. This is because every time a function is called, a new environment is created to host excecution. A function has no way to tell what happened last time it ran.

### dynamic lookup

Lexical scoping determines WHERE to look for functions, not WHEN. R looks for values when function is run, not creates. This means the output of the function can be different depending on the objects outside the environment

```{r}
f <- function() x
x <- 15
f()

x <- 20
f()
```

* Generally want to avoid this behaviour because it means the function is not self contained.

### Every operation is a function call!!

* including infix opertors (like + - etc), control flow (if,else etc), and even the curly brace {!

```{r}
x <- 10; y <- 5
# get the same answer:
x + y
`+`(x, y)
```

```{r}
# is often useful to treat secial functions like ordinary functions eg
add <- function(x, y) x + y
sapply(1:10, add, 3)
sapply(1:10, `+`, 3)
```

### Function arguments

* useful to ditiinguish between actual arguments and formal arguments

```{r}
# calling functions

f <- function(abcdef, bcde1, bcde2){
  list(a = abcdef, b1 = bcde1, b2 = bcde2)
}

#positional matching
str(f(1, 2, 3))

str(f(2, 3, abcdef = 1))

# can abbreviate
str(f(2, 3, a = 1))

#cant to this as not clear
#str(f(1, 3, b = 2))
```

* dont be stupid, make sure your function calls are concise and clear in meaning

### Calling a function from a list of arguments

```{r}
args <- list(1:10, na.rm = TRUE)

# apply a list of function arguments to a function using do.call()
do.call(mean, args)
# same as:
mean(1:10, na.rm = TRUE)
```

### default and missing values

* functions can have default values eg function(a = 10, b = 2)
* as R arguents evaluated lazily, the default values can be defined in terms of others eg function(a = 10, b = a * 2)
* default arguments can be defined in terms of variables in the function but is arguably bad practice.
* use the missing() function to see if argument was supplied or not 

```{r}
i <- function(a, b) {
  c(missing(a), missing(b))
}
i()
i(a = 1)
i(b = 1)
i(1, 2)
```

### Lazy evaluation

* R arguments are evaluated lazily- they are only evaluated if they are actually used.
```{r}
f <- function(x){
  10
}
f(stop('ERROR'))

# have to force to evaluate
f <- function(x){
#  force(x)
  10
}
f(stop('ERROR'))

```

* this is important, for example:

```{r}
add <- function(x){
  function(y) x +y
}

adders <- lapply(1:10, add)
adders[[1]](10)
adders[[10]](10)

# as the loop is complete, final value of x is 10. Therefore will add 10 to the input, regardless
# so manually force evaluation

add <- function(x){
  force(x)
  function(y) x +y
}

adders <- lapply(1:10, add)
adders[[1]](10)
adders[[10]](10)
```

* laziness useful in if statements- the second statement beow is only evaluated if the first is true. If it wasnt the statemnet would return and error because NULL >0 is a logical vector of length zero and not a valid input to if

```{r}
x <- NULL
if (!is.null(x) && x > 0){ }
```

### ...

* ... matches any arguments not otherwise matched, and can be easily passed to other functions. Useful if you want to collect arguments to other functions. Makes functions very flexible, but sometimes hard to understand how to use. Also, is not spell checked, so errors can slip in.


### Infix functions

* most functions in R are prefix functions, however can also create and use infix functions. Built in ones eg %in% %/% as well as + - = etc etc

```{r}
# build an infix operator that pastes together strings
`%+%` <- function(a, b) paste(a, b, sep = "" )
"new" %+% " string"
`%+%`("new", " string")

# a useful infix function: provides a default if output of statement is NULL
`%||%` <- function(a, b) if(!is.null(a)) a else b
# function_that_might_return_null() %||% default value
```

### Replacement Functions

* replacement functions act as if they modify their arguments in place

```{r}
`second<-` <- function(x, value){
  x[2] <- value
  x
}
x <- 1:10
second(x) <- 5L
x
```

* often useful to combine replacement and subsetting

```{r}
x <- c(a = 1, b = 2, c = 3)
names(x)
names(x)[2] = "two"
names(x)

# this works because names(x)[2] <- "two" is written as if you wrote
# `*tmp*` <- names(x)
# `*tmp*`[2] <- "two"
# names(x) <- `*two*`
```

### Return Values

* last expression evaluated is returned. Good practice only to use return() if specifying an error check where a function returns early 

* R functions (generally) have copy-on-modify semantics- modifyig the function argument does not change the origional value

* functions can return invisible values
```{r}
f1 <- function() invisible(1)
f1()
(f1())

# assignment function uses invisible, eg
a <- 2
(a <- 2)
```

### On exit

* functions can set up triggers to occur when function finishes. code in on.exit() runs regardless of success of function

```{r}
in_dir <- function(dir, code){
  old <- setwd(dir)
  on.exit(setwd(old))
  
  force(code)
  
}
getwd()
in_dir("~", getwd())
```
